---
output: 
  html_document: 
    df_print: default
---

# Homework 1: Data Visualization & Google Trends on Three Different Measures from Turkey

In this homework, time series analysis of interest rate levels, unemployment rate levels, and percentage change of cost of living indices for wage earners in İstanbul will be performed over the years using monthly data. The aim is to analyze specific characteristics of these chosen datasets and show that there is a pattern both with regard to time and each other. In addition, GoogleTrends data for search keywords "iskur" and "zam" are included to deepen the analysis.

## 1) Data Gathering and Manipulations

### a. Monthly Interest Rates of Turkey

This measure's data set consists of monthly reported values of interest rates in Turkey. The data is taken between 2002 and 2021.After the necessary manipulations and some cleaning, head of this data frame is as follows:

```{r echo=FALSE, include=TRUE, warning=FALSE}
#Necessary libraries, please install them beforehand if they are not already installed.
library(xlsx)
library(lubridate)
library(ggplot2)
library(zoo)
require(GGally)
setwd("~/R")
InterestEVDS=read.xlsx("EVDS-Interest Rate.xlsx",sheetIndex=1, header=TRUE)
InterestEVDS = InterestEVDS[1:240,1:2]
InterestEVDS[,"Date"]=as.Date(as.yearmon(InterestEVDS[,"Date"],format="%Y-%m"))
colnames(InterestEVDS) <- c("Date","InterestRates")
InterestEVDS[,"InterestRates"]=as.numeric(InterestEVDS[,"InterestRates"])
head(InterestEVDS)
```

Dummy day information to the date is added since it is easier to plot and visualize the data with the full date.

### b. Monthly Unemployment Rates of Turkey

This second measure's data set consists of monthly reported values of unemployment rates as percentage in Turkey. The data is taken between 2014 and 2021 as it was only available for this period .After the necessary manipulations and some cleaning, head of this data frame is as follows:

```{r echo=FALSE, include=TRUE, warning=FALSE}
setwd("~/R")
UnemploymentEVDS=read.xlsx("EVDS-Unemployment Rate.xlsx",sheetIndex=1, header=TRUE)
UnemploymentEVDS = UnemploymentEVDS[1:96,1:2]
UnemploymentEVDS[,"Date"]=as.Date(as.yearmon(UnemploymentEVDS[,"Date"],format="%Y-%m"))
colnames(UnemploymentEVDS) <- c("Date","UnemploymentRates")
UnemploymentEVDS[,"UnemploymentRates"]=as.numeric(UnemploymentEVDS[,"UnemploymentRates"])
head(UnemploymentEVDS)
```

### c. Monthly Percentage Change of Cost of Living Indices for Wage Earners-Istanbul

This last dataset covers the information about general expenses of a typical wage earner in İstanbul. Since first two data are in the form of rates, this measure is chosen as percentage change instead of gross levels. In substance, this data mimics CPI for İstanbul.Performing the same manipulations as above:

```{r echo=FALSE, include=TRUE, warning=FALSE}
setwd("~/R")
COLEVDS=read.xlsx("EVDS-Cost of Living.xlsx",sheetIndex=1, header=TRUE)
COLEVDS = COLEVDS[1:240,1:2]
COLEVDS[,"Date"]=as.Date(as.yearmon(COLEVDS[,"Date"],format="%Y-%m"))
colnames(COLEVDS) <- c("Date","CostOfLiving")
COLEVDS[,"CostOfLiving"]=as.numeric(COLEVDS[,"CostOfLiving"])
head(COLEVDS)

```

### d. Google Trends Search for "iskur" & "zam"

As Google Trends search keywords "iskur" and "zam" are selected. "zam" means "increase in prices/wages" in Turkish. It is a widely used word in internet searches when people are expecting a raise in minimum wage. "iskur" is the legal regulation regarding public employment services in Turkey and people wanting to be employed most probably visits the iskur's website. The data provided by Google Trends for this keyword was also monthly.

```{r echo=FALSE, include=TRUE, warning=FALSE}
setwd("~/R")
trends=read.csv("multiTimeline.csv", header=TRUE)
trends[,"Ay"]=as.Date(as.yearmon(trends[,"Ay"],format="%Y-%m"))
colnames(trends) <- c("Date","işkur","zam")
trends[,"işkur"]=as.numeric(trends[,"işkur"])
trends[,"zam"]=as.numeric(trends[,"zam"])
head(trends)
```

## 2) Data Visualizations

### a. Data from EVDS

These datasets are merged into one dataframe containing only the years that do not have NA values for any column (2014-2021).

```{r}
EVDS1 <- merge(InterestEVDS,UnemploymentEVDS,by='Date')
EVDS <- merge(EVDS1,COLEVDS,by='Date')

ggplot(EVDS, aes(Date,InterestRates)) +
  geom_line(na.rm=TRUE,colour='red')+theme_classic()+scale_x_date(date_breaks = "1 year", date_labels = "%Y %b", date_minor_breaks = "1 month")
```

If we look at the plot of Interest Rates over time, we can see fluctuations over time and general trend to increase except for the COVID period in March 2020. We are also able to observe seemingly unusual peak in close to the beginning of 2019. This is due to volatility in Turkey's economy and politics.

```{r}
ggplot(EVDS, aes(Date,UnemploymentRates)) +
  geom_line(na.rm=TRUE,colour='blue')+theme_classic()+scale_x_date(date_breaks = "1 year", date_labels = "%Y %b", date_minor_breaks = "1 month")
```

Even though, fluctuations seems more than interest rates this is due to scale differences and we see again a general increase in unemployment rates. The effect of COVID is not as obvious as it was in the previous case. Seasonal workers' effect is also observed.

```{r}
ggplot(EVDS, aes(Date,CostOfLiving)) +
  geom_line(na.rm=TRUE,colour='green')+scale_x_date(date_breaks = "1 year", date_labels = "%Y %b", date_minor_breaks = "1 month")+theme_classic()
```

Lastly, change in the Cost of Living Indices for İstanbul is plotted over time and it is observed that most of the values are larger than 0, meaning the life in İstanbul tends to get more and more expensive over time. Here, the most interesting part is the 2022 January and how the increase is almost 10% just in one month.

```{r}
ggplot(EVDS,aes(x=factor(month(Date)),y=InterestRates))+geom_bar(stat="identity",aes(fill=month(Date)))+facet_wrap(~year(Date))+
  theme(legend.position = "none",
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank())+
  labs(x="Month",title="Histograms of Monthly Inflation Rates of Different Years")+scale_fill_distiller(palette="Pastel2")

ggplot(EVDS,aes(x=factor(month(Date)),y=UnemploymentRates))+geom_bar(stat="identity",aes(fill=month(Date)))+facet_wrap(~year(Date))+
  theme(legend.position = "none",
        axis.ticks.x = element_blank(),
       axis.text.x = element_blank())+
  labs(x="Month",title="Histograms of Monthly Unemployment Rates of Different Years")+scale_fill_distiller(palette="Accent")
ggplot(EVDS,aes(x=factor(month(Date)),y=CostOfLiving))+geom_bar(stat="identity",aes(fill=month(Date)))+facet_wrap(~year(Date))+
  theme(legend.position = "none",
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank())+
  labs(x="Month",title="Histograms of Monthly Change in Cost of Living in İstanbul for Different Years")+scale_fill_distiller(palette="Set2")

```

We can then try to see the effect of months in each statistics by looking at their histograms over months. We can observe that cost of living changes more impulsively compared to the other two which are pretty much at constant level throughout the year.

### b. Data from Google Trends

```{r echo=FALSE, include=TRUE, warning=FALSE}
ggplot(data=trends,aes(x=Date))+geom_line(aes(y=iskur))+geom_line(aes(y=zam),color="red")+theme_classic()+labs(y=" ")
ccf(trends$zam,trends$iskur,lag.max=1000, plot=TRUE, main="Search for 'işkur' & 'zam'")  
```

It is shown that there is again a general increase in the search for both "iskur" and "zam" with some fluctuations. Also,there may be a relation between the search for "iskur" and "zam" however there seems to be lag that makes it hard to see the correlation clearly. To get rid of the effects of this, the lag that gives the biggest correlation is found and the plot is redone considering this lag:

```{r echo=FALSE, include=TRUE, warning=FALSE}
Find_Abs_Max_CCF<- function(a,b)
{
  d <- ccf(a, b, plot = FALSE, lag.max = length(a)-5)
  cor = d$acf[,,1]
  abscor = abs(d$acf[,,1])
  lag = d$lag[,,1]
  res = data.frame(cor,lag)
  absres = data.frame(abscor,lag)
  absres_max = res[which.max(absres$abscor),]
  return(absres_max)
}
Find_Abs_Max_CCF(trends$zam,trends$iskur)
ggplot(data=trends,aes(x=Date))+geom_line(aes(y=c(tail(iskur,-21),rep(NA,21))))+geom_line(aes(y=zam),color="red")

```

### c. Data from both Google Trends and EVDS

Lastly, I wanted to plot unemployment rate versus "iskur" search in time as strong correlation between them is expected.

```{r echo=FALSE, include=TRUE, warning=FALSE}
ggplot()+geom_line(data=trends[121:216,1:3],aes(x=Date,y=iskur), color="yellow")+geom_line(data=EVDS, aes(x=Date, y=UnemploymentRates),color="pink")+theme_dark()+labs(y="")
cor(trends[121:216,1:3]$iskur,EVDS$UnemploymentRates)

```

High correlation around 0.59 is observed as expected.

## 3) Correlations and Comparisons

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
ggpairs(EVDS)
```

From the results, it can be seen that there is a strong correlation between pairs of:

-   Time- Unemployment Rate

-   Time - Interest Rates

-   Interest Rates- Unemployment Rate

    However, percentage change in the cost of living indices for İstanbul does not seem to be correlated with the other two.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
ggpairs(trends)

```

All pairs are strongly correlated. Also the relation of "zam" to time looks logarithmic which suggests model may be open to improvements. Similar "linear" increase in unemployment rate is observed in the search for "iskur"

## 4) Conclusion

The aim of this paper was to analyze the relationship of chosen datasets with time. Concluding remarks are as following:

-   Increase in unemployment rates and search for "iskur" on Google is highly correlated as expected.

-    Cost of Living in İstanbul for a typical wage earner gets higher and higher with time, gaining a sharp acceleration lately.

-   There is a strong correlation between unemployment rates and interest rates

-   There is lag between search for "zam" and "iskur" in GoogleTrends data.

For R file click [here.](https://github.com/BU-IE-360/spring22-elifatilgan/blob/gh-pages/files/hw1.R)
